---
title: "Boston Housing"
author: "Tristan-Mihai Radulescu, Hugo Ponthieu, Benoit Planche"
output:
  html_document:
    df_print: paged
---

# **Introduction**
This report analyzes the `BostonHousing` dataset, which contains information about housing in Boston. We will explore the relationships between several variables and the median home value (`medv`).  

---

```{r setup, include=FALSE}
# Load necessary libraries
library(ggplot2)
library(knitr)
library(tidyverse)
library(broom)
library(GGally)
library(MASS)
library(randomForest)
library(rsample)
library(dplyr)
library(ipred)
library(caret)
library(rpart)
library(rpart.plot)
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
# Load the dataset
BostonHousing <- read.csv("./BostonHousing.csv")

# Display first rows
head(BostonHousing)

# Structure of the dataset
# str(BostonHousing)
```


## **1. Distribution of Key Variables**
We analyze the distribution of home prices (`medv`).  

```{r valeur-mediane, echo=FALSE}
ggplot(BostonHousing, aes(x = medv)) + 
  geom_histogram(fill = "blue", bins = 30, alpha = 0.7) +
  labs(title = "Distribution of Home Prices",
       x = "MEDV (in $1000s)", 
       y = "Frequency")
```

---

## **2. Correlations Between Variables**
We compute the correlation matrix and display values associated with `medv`.  

```{r}
# Compute correlation matrix
cor_matrix <- cor(BostonHousing)

# Display correlations with MEDV
cor_matrix["medv", ]
```

**Observations**:  
- **Positive correlation** with `rm` (number of rooms).  
- **Negative correlation** with `lstat` (percentage of lower-status population).  
- **`tax` and `crim` have a negative relationship** with `medv`.  


### **2.1 Pairs**

```{r pairs, echo=FALSE}
pairs(BostonHousing)
```

---

### **2.2 Key Relationships**

```{r nb-piece, echo=FALSE}
# Relationship between RM (number of rooms) and MEDV
ggplot(BostonHousing, aes(x = rm, y = medv)) +
  geom_point(alpha = 0.6, color = "blue") +
  geom_smooth(method = "lm", col = "red") +
  labs(title = "Relationship Between RM and MEDV", x = "Average Number of Rooms", y = "Median Home Value (in $1000s)")
```

```{r pop-pauvre, echo=FALSE}
# Relationship between LSTAT (% lower-status population) and MEDV
ggplot(BostonHousing, aes(x = lstat, y = medv)) +
  geom_point(alpha = 0.6, color = "purple") +
  geom_smooth(method = "lm", col = "red") +
  labs(title = "Relationship Between LSTAT and MEDV", x = "Lower-Status Population (%)", y = "Median Home Value (in $1000s)")
```

---

## **3. Impact of Crime and Taxation on Home Prices**
### **3.1 Distribution of `crim` and `tax`**
```{r tax-crim, echo=FALSE}
ggplot(BostonHousing, aes(x = crim)) +
  geom_histogram(fill = "red", bins = 30, alpha = 0.7) +
  labs(title = "Distribution of Crime Rate", x = "CRIM", y = "Frequency") +
  scale_x_log10()

ggplot(BostonHousing, aes(x = tax)) +
  geom_histogram(fill = "blue", bins = 30, alpha = 0.7) +
  labs(title = "Distribution of Tax Rate", x = "TAX", y = "Frequency")
```

**Observations**:  
- **`crim` is highly skewed**: most neighborhoods have low crime rates, but some areas have extreme values.  
- **`tax` shows distinct peaks**, suggesting fixed tax rates in certain areas.  

---

### **3.2 Relationship Between Crime Rate and Home Prices**
```{r tax-crim2, echo=FALSE}
ggplot(BostonHousing, aes(x = crim, y = medv)) +
  geom_point(alpha = 0.5, color = "red") +
  geom_smooth(method = "lm", col = "black") +
  scale_x_log10() +
  labs(title = "Relationship Between CRIM and MEDV",
       x = "Crime Rate (log)", 
       y = "Median Home Value (in $1000s)")
```

**Observation**: Home prices **decrease as crime rates increase**.  

---

## **3.3 Relationship Between Taxation and Home Prices**
```{r tax-crim3, echo=FALSE}
ggplot(BostonHousing, aes(x = tax, y = medv)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", col = "black") +
  labs(title = "Relationship Between TAX and MEDV",
       x = "Tax Rate", 
       y = "Median Home Value (in $1000s)")
```

---

**Observation**: A **higher tax rate is associated with lower home prices**.  

```{r reg-multiple, echo=FALSE}
reg_multi <- lm(medv ~ crim + tax, data = BostonHousing)
summary(reg_multi)
```

**Results**:  
- **The coefficients for `crim` and `tax` are negative**, confirming their negative impact on `medv`.  
- **p-values < 0.05** indicate these effects are statistically significant.  

---

## 4. Linear regressions

### 4.1 Based on pairs

```{r}
lin_reg <- lm(medv ~ rm, data = BostonHousing)
ggplot(data = BostonHousing, aes(x = medv, y = rm)) +
  geom_point() +
  geom_smooth(method = "lm")
```


### 4.2 Backward selection

```{r}
model <- lm(medv ~ ., data = BostonHousing)

summary(model)

backward_model <- step(model, direction = "backward", trace = FALSE)

summary(backward_model)


plot(backward_model)
```

### 4.3 Multiple linear regression with normalized data and backward selection

```{r}
normalized_data <- preProcess(BostonHousing, method = c("center", "scale"))

normalized_data <- predict(normalized_data, BostonHousing)
```
#### 4.3.1 Boxplots

```{r}
boxplot(BostonHousing)
boxplot(normalized_data)
```

#### 4.3.2 Modelization

```{r}
model <- lm(medv ~ ., data = normalized_data)
summary(model)
backward_model <- step(model, direction = "backward", trace = FALSE)
summary(backward_model)
plot(backward_model)
```

---

### 4.4. Forward selection
#### 4.4.1. Sampling data
```{r}
# Load necessary libraries
set.seed(123)
sample_size <- floor(0.8 * nrow(BostonHousing))
train_indices <- sample(seq_len(nrow(BostonHousing)), size = sample_size)
train_data <- BostonHousing[train_indices, ]
test_data <- BostonHousing[-train_indices, ]

initial_model <- lm(medv ~ 1, data = train_data)


```

#### 4.4.2. Applying the algorithm

```{r}
forward_model <- stepAIC(initial_model, direction = "forward", scope = list(lower=initial_model, upper=~crim+ zn+ indus+chas +nox+rm+ age +dis+rad+tax+ptratio+b+lstat ))
```



```{r}
plot(forward_model, which=2)
summary(forward_model)

test_predictions <- predict(forward_model, newdata = test_data)

actual_vs_predicted <- data.frame(Actual = test_data$medv, Predicted = test_predictions)

residuals <- data.frame(Predicted = test_predictions, Residuals = test_data$medv - test_predictions)

ggplot(residuals, aes(x = Predicted, y = Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residual Plot", x = "Predicted Values", y = "Residuals") +
  theme_minimal() +
  # Ensure the full range of predicted values is shown
  coord_cartesian(xlim = range(test_predictions))
correlation_matrix <- cor(train_data)
```

#### 4.4.3. Correlation Matrix
```{r}
ggplot(reshape2::melt(correlation_matrix), aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1, 1), space = "Lab", name = "Pearson\nCorrelation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 12, hjust = 1)) +
  coord_fixed()

```

## 5. Random forest model
```{r}
set.seed(123)
rf_model <- randomForest(medv ~ ., data = train_data, importance = TRUE)
rf_predictions <- predict(rf_model, newdata = test_data)
# Evaluate the model
actual_vs_predicted_rf <- data.frame(Actual = test_data$medv, Predicted = rf_predictions)

# Plot actual vs predicted values
ggplot(actual_vs_predicted_rf, aes(x = Actual, y = Predicted)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Actual vs Predicted Values (Random Forest)", x = "Actual Values", y = "Predicted Values") +
  theme_minimal()

```

### 5.1. Decision Tree
```{r}
# Plot a tree using rpart
tree_model <- rpart(medv ~ ., data = train_data)

rpart.plot(tree_model, cex = 0.8)
```
As we can see, the model is not very good. We can try to improve it by using a tree. Intuitively, we can see that our data is segmented in three groups. One group is the one with the lowest medv, one group is the one with the highest medv and the third group is in between. But let's use trees to segment these groups correctly.
