---
title: "Boston Housing"
author: "Tristan-Mihai Radulescu, Hugo Ponthieu, Benoit Planche"
output:
  html_document:
    df_print: paged
---

# **Introduction**
This report analyzes the `BostonHousing` dataset, which contains information about housing in Boston. We will explore the relationships between several variables and the median home value (`medv`).  

---

# **1. Loading Libraries and Data**
```{r setup, include=FALSE}
# Load necessary libraries
library(ggplot2)
library(knitr)
library(tidyverse)
library(broom)
library(GGally)
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
# Load the dataset
BostonHousing <- read.csv("./BostonHousing.csv")

# Display first rows
head(BostonHousing)

# Structure of the dataset
str(BostonHousing)
```

## Histogram on Medv

```{r}
hist(BostonHousing$medv, breaks = 15)
```

On this histogram, we can see that the distribution is not normal. The mode is at arout 20 to 25.


## Linear regression

```{r}
pairs(
  data = BostonHousing,
  ~.
)
```

```{r}
lin_reg <- lm(medv ~ rm, data = BostonHousing)
ggplot(data = BostonHousing, aes(x = medv, y = rm)) +
  geom_point() +
  geom_smooth(method = "lm")
```

## Linear Multiple Regression

### Backward Elimination

```{r}
model <- lm(medv ~ ., data = BostonHousing)

summary(model)

backward_model <- step(model, direction = "backward", trace = FALSE)

summary(backward_model)


plot(backward_model)
```

## Perfect linear regression

```{r}
lin_reg <- lm(medv ~ medv, data = BostonHousing)
plot(lin_reg)
```

## Worst linear regression

```{r}
lin_reg <- lm(medv ~ indus, data = BostonHousing)
plot(lin_reg)
```
Can you make a box plot of the value of the rivers in the BostonHousing dataset? 

```{r}
# Load necessary libraries
library(ggplot2)

# Create a boxplot for 'medv' colored by 'chas'
ggplot(BostonHousing, aes(x = factor(chas), y = medv, fill = factor(chas))) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Boxplot of Median Home Values by River Proximity",
       x = "Proximity to the River (0 = No, 1 = Yes)",
       y = "Median Home Value (in $1000s)",
       fill = "Proximity") +
  scale_fill_manual(values = c("0" = "red", "1" = "blue")) +
  theme_minimal()
```

**Observation**: Proximity to the river **seems to have a slight impact** on home prices, but the distribution remains fairly similar.  

---

# **5. Distribution of Key Variables**
We analyze the distribution of home prices (`medv`).  

```{r valeur-mediane, echo=FALSE}
ggplot(BostonHousing, aes(x = medv)) + 
  geom_histogram(fill = "blue", bins = 30, alpha = 0.7) +
  labs(title = "Distribution of Home Prices",
       x = "MEDV (in $1000s)", 
       y = "Frequency")
```

---

# **6. Correlations Between Variables**
We compute the correlation matrix and display values associated with `medv`.  

```{r}
# Compute correlation matrix
cor_matrix <- cor(BostonHousing)

# Display correlations with MEDV
cor_matrix["medv", ]
```

**Observations**:  
- **Positive correlation** with `rm` (number of rooms).  
- **Negative correlation** with `lstat` (percentage of lower-status population).  
- **`tax` and `crim` have a negative relationship** with `medv`.  

---

# **6.1 Key Relationships**
```{r nb-piece, echo=FALSE}
# Relationship between RM (number of rooms) and MEDV
ggplot(BostonHousing, aes(x = rm, y = medv)) +
  geom_point(alpha = 0.6, color = "blue") +
  geom_smooth(method = "lm", col = "red") +
  labs(title = "Relationship Between RM and MEDV", x = "Average Number of Rooms", y = "Median Home Value (in $1000s)")
```

```{r pop-pauvre, echo=FALSE}
# Relationship between LSTAT (% lower-status population) and MEDV
ggplot(BostonHousing, aes(x = lstat, y = medv)) +
  geom_point(alpha = 0.6, color = "purple") +
  geom_smooth(method = "lm", col = "red") +
  labs(title = "Relationship Between LSTAT and MEDV", x = "Lower-Status Population (%)", y = "Median Home Value (in $1000s)")
```

---

# ðŸ  **7. Impact of Crime and Taxation on Home Prices**
## ðŸ” **7.1 Distribution of `crim` and `tax`**
```{r tax-crim, echo=FALSE}
ggplot(BostonHousing, aes(x = crim)) +
  geom_histogram(fill = "red", bins = 30, alpha = 0.7) +
  labs(title = "Distribution of Crime Rate", x = "CRIM", y = "Frequency") +
  scale_x_log10()

ggplot(BostonHousing, aes(x = tax)) +
  geom_histogram(fill = "blue", bins = 30, alpha = 0.7) +
  labs(title = "Distribution of Tax Rate", x = "TAX", y = "Frequency")
```

**Observations**:  
- **`crim` is highly skewed**: most neighborhoods have low crime rates, but some areas have extreme values.  
- **`tax` shows distinct peaks**, suggesting fixed tax rates in certain areas.  

---

## **7.2 Relationship Between Crime Rate and Home Prices**
```{r tax-crim2, echo=FALSE}
ggplot(BostonHousing, aes(x = crim, y = medv)) +
  geom_point(alpha = 0.5, color = "red") +
  geom_smooth(method = "lm", col = "black") +
  scale_x_log10() +
  labs(title = "Relationship Between CRIM and MEDV",
       x = "Crime Rate (log)", 
       y = "Median Home Value (in $1000s)")
```

**Observation**: Home prices **decrease as crime rates increase**.  

---

## **7.3 Relationship Between Taxation and Home Prices**
```{r tax-crim3, echo=FALSE}
ggplot(BostonHousing, aes(x = tax, y = medv)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", col = "black") +
  labs(title = "Relationship Between TAX and MEDV",
       x = "Tax Rate", 
       y = "Median Home Value (in $1000s)")
```

**Observation**: A **higher tax rate is associated with lower home prices**.  

---

# **8. Multiple Regression Analysis**
```{r reg-multiple, echo=FALSE}
reg_multi <- lm(medv ~ crim + tax, data = BostonHousing)
summary(reg_multi)
```

**Results**:  
- **The coefficients for `crim` and `tax` are negative**, confirming their negative impact on `medv`.  
- **p-values < 0.05** indicate these effects are statistically significant.  

---

# **Conclusion**
This analysis shows that **the number of rooms (`rm`), crime rate (`crim`), and tax rate (`tax`) strongly influence home prices**.  

**Suggestions:** Further testing with advanced multiple regression models and interaction effects. ðŸš€  
```